# Subsetting Data Structures {#subset}

We will create the same data structures in section \@ref(structures) of the notes to use as examples in this section.

```{r}
x <- c(5,10,15,20,25)
y <- c(3,6,9,12,15)
z <- c(x,y)
log_vect <- c(TRUE, FALSE, T, F)
char_vect <- c("ByOne", "ByTwo", "ByThree", "ByFour", "ByFive")
mult_mat <- cbind(char_vect,x,y)
mult_data <- data.frame(char_vect,x,y)
mult_list <- list(mult_data, z, log_vect, 7)
```

## Overview

Once a data structure is created and stored, it may be useful in subsequent analyses to subset that data structure. You will learn how to subset data structures using base R functions in this section. Similar operations can be performed using functions from the `dplyr` package, and you will learn these in a later section. 

## Subsetting with Brackets

### Subsetting Vectors

The typical notation to subset vectors is a set of brackets used directly after the variable name of the vector. The information given inside the brackets specifies which values should be extracted from the vector. 

The position index of any vector begins at 1 and uses consecutive positive integers up to the length of the vector. For example, to obtain the 3rd element of the vector `x`:

```{r}
x[3]
x ## for comparison
```

Note: Indexing in R starts from 1. R is designed for data analysis, statistics, and mathematics, where standard notations begin counting at 1 (e.g. statisticians think in terms of sample size). This is unlike indexing starting from 0 for other software such as Python and C, who audiences tend to be programmers and software engineers.

Multiple positions can be specified by using a vector in the brackets.

```{r}
x[c(1,3)] ##1st and 3rd positions
x[2:3] ##2nd and 3rd posistions
```

If a position is duplicated in the brackets, the corresponding value in the vector will be subset twice. 

```{r}
x[c(1,1)] ##1st position displayed twice
```

If a negative sign is used in the brackets, the specified value(s) will be omitted instead of included in the subset.

```{r}
x[-1] ##omit position 1
x[-(2:3)] ##omit positions 2, 3
x[-c(1,4)] ##omit positions 1, 4
```

Instead of specifying the positions that you would like included or excluded from the subset, you can specify a logical vector in the brackets. R will include the elements corresponding to TRUE and exclude the elements corresponding to FALSE.  

```{r}
x[c(F,T,T,F,T)]
```

If the number of logical inputs does not match the length of the vector, R will repeat the logical inputs until there is a logical input corresponding to each vector element.

```{r}
x[c(F,T)]
```

The resulting subset vectors will only be stored if you define a variable name and assign the subset vector to it.

```{r}
evens <- x[c(F,T)]
x
evens
```

The brackets following a vector can also be used to replace or add one or more values in a vector.

```{r}
x2 <- x

x2[6] <- 30 ##place 30 in position 6
x2

x2[7:9] <- c(35,40,45) ##then add 3 values
x2

x2[c(1,3)] <- 0 ##change value of position 1, 3 to 0
x2
```

### Subsetting Matrices and Data Frames 

To subset a matrix or data frame, a similar bracket notation is used. However, because matrices and data frames are two-dimensional structures, two arguments are required in the brackets. The first argument references the row index and the second input references the column index.

Note: subsetting data frames is very common when analyzing data, as data are usually stored in data frames.

The examples below are applied to a data frame. Subsetting matrices work in the same way.

```{r}
mult_data

## Subsetting the value in the first row and second column
mult_data[1,2]

## Subsetting the values in the first two rows and first two columns
mult_data[1:2,1:2]

## Subsetting the values in rows 1, 4 and columns 2, 3
mult_data[c(1,4),2:3]

## Subsetting the entire first row.
## Comma as the 2nd argument means want all columns
mult_data[1,]

## Subsetting the entire second column
## Comma as the 1st argument means want all rows
mult_data[,2]

## Subsetting all but the first row
mult_data[-1,]

## Subsetting all but the third column
mult_data[,-3]

## Subsetting the third column twice
mult_data[,c(3,3)]

## Subsetting columns 1, 2, 3 and duplicating the third column
mult_data[,c(1:3,3)]
```

### Subsetting Lists

A list is subset to any of its elements by using double brackets, `[[ ]]`.

```{r}
mult_list[[3]]
```
Double brackets are limited to one value. In order to subset multiple elements of a list, single brackets can be used. 

```{r}
mult_list[2:3]
```
### 







