# Subsetting Data Structures {#subset}

We will create the same data structures in section \@ref(structures) of the notes to use as examples in this section.

```{r}
x <- c(5,10,15,20,25)
y <- c(3,6,9,12,15)
z <- c(x,y)
log_vect <- c(TRUE, FALSE, T, F)
char_vect <- c("ByOne", "ByTwo", "ByThree", "ByFour", "ByFive")
mult_mat <- cbind(char_vect,x,y)
mult_data <- data.frame(char_vect,x,y)
mult_list <- list(mult_data, z, log_vect, 7)
```

## Overview

Once a data structure is created and stored, it may be useful in subsequent analyses to subset that data structure. You will learn how to subset data structures using base R functions in this section. Similar operations can be performed using functions from the `dplyr` package, and you will learn these in a later section. 

## Subsetting Vectors

The typical notation to subset vectors is a set of brackets used directly after the variable name of the vector. The information given inside the brackets specifies which values should be extracted from the vector. 

The position index of any vector begins at 1 and uses consecutive positive integers up to the length of the vector. For example, to obtain the 3rd element of the vector `x`:

```{r}
x[3]
x ## for comparison
```

Note: Indexing in R starts from 1. R is designed for data analysis, statistics, and mathematics, where standard notations begin counting at 1 (e.g. statisticians think in terms of sample size). This is unlike indexing starting from 0 for other software such as Python and C, who audiences tend to be programmers and software engineers.

Multiple positions can be specified by using a vector in the brackets.

```{r}
x[c(1,3)] ##1st and 3rd positions
x[2:3] ##2nd and 3rd posistions
```

If a position is duplicated in the brackets, the corresponding value in the vector will be subset twice. 

```{r}
x[c(1,1)] ##1st position displayed twice
```

If a negative sign is used in the brackets, the specified value(s) will be omitted instead of included in the subset.

```{r}
x[-1] ##omit position 1
x[-(2:3)] ##omit positions 2, 3
x[-c(1,4)] ##omit positions 1, 4
```

Instead of specifying the positions that you would like included or excluded from the subset, you can specify a logical vector in the brackets. R will include the elements corresponding to TRUE and exclude the elements corresponding to FALSE.  

```{r}
x[c(F,T,T,F,T)]
```

If the number of logical inputs does not match the length of the vector, R will repeat the logical inputs until there is a logical input corresponding to each vector element.

```{r}
x[c(F,T)]
```

The resulting subset vectors will only be stored if you define a variable name and assign the subset vector to it.

```{r}
evens <- x[c(F,T)]
x
evens
```

The brackets following a vector can also be used to replace or add one or more values in a vector.

```{r}
x2 <- x

x2[6] <- 30 ##place 30 in position 6
x2

x2[7:9] <- c(35,40,45) 
x2

x2[c(1,3)] <- 0 ##change value of position 1, 3 to 0
x2
```

## Subsetting Data Frames and Lists

