# Working with Data Structures {#subset}

We will create the same data structures in section \@ref(structures) of the notes to use as examples in this section.

```{r}
x <- c(5,10,15,20,25)
y <- c(3,6,9,12,15)
z <- c(x,y)
log_vect <- c(TRUE, FALSE, T, F)
char_vect <- c("ByOne", "ByTwo", "ByThree", "ByFour", "ByFive")
mult_mat <- cbind(char_vect,x,y)
mult_data <- data.frame(char_vect,x,y)
mult_list <- list(mult_data, z, log_vect, 7)
```

## Subsetting Data Structures

Once a data structure is created and stored, it may be useful in subsequent analyses to subset that data structure. You will learn how to subset data structures using base R functions in this section. Similar operations can be performed using functions from the `dplyr` package, and you will learn these in a later section. 

### Subsetting with Brackets

#### Subsetting Vectors

The typical notation to subset vectors is a set of brackets used directly after the variable name of the vector. The information given inside the brackets specifies which values should be extracted from the vector. 

The position index of any vector begins at 1 and uses consecutive positive integers up to the length of the vector. For example, to obtain the 3rd element of the vector `x`:

```{r}
x[3]
x ## for comparison
```

Note: Indexing in R starts from 1. R is designed for data analysis, statistics, and mathematics, where standard notations begin counting at 1 (e.g. statisticians think in terms of sample size). This is unlike indexing starting from 0 for other software such as Python and C, who audiences tend to be programmers and software engineers.

Multiple positions can be specified by using a vector in the brackets.

```{r}
x[c(1,3)] ##1st and 3rd positions
x[2:3] ##2nd and 3rd posistions
```

If a position is duplicated in the brackets, the corresponding value in the vector will be subset twice. 

```{r}
x[c(1,1)] ##1st position displayed twice
```

If a negative sign is used in the brackets, the specified value(s) will be omitted instead of included in the subset.

```{r}
x[-1] ##omit position 1
x[-(2:3)] ##omit positions 2, 3
x[-c(1,4)] ##omit positions 1, 4
```

Instead of specifying the positions that you would like included or excluded from the subset, you can specify a logical vector in the brackets. R will include the elements corresponding to TRUE and exclude the elements corresponding to FALSE.  

```{r}
x[c(F,T,T,F,T)]
```

If the number of logical inputs does not match the length of the vector, R will repeat the logical inputs until there is a logical input corresponding to each vector element.

```{r}
x[c(F,T)]
```

The resulting subset vectors will only be stored if you define a variable name and assign the subset vector to it.

```{r}
evens <- x[c(F,T)]
x
evens
```

The brackets following a vector can also be used to replace or add one or more values in a vector.

```{r}
x2 <- x

x2[6] <- 30 ##place 30 in position 6
x2

x2[7:9] <- c(35,40,45) ##then add 3 values
x2

x2[c(1,3)] <- 0 ##change value of position 1, 3 to 0
x2
```

#### Subsetting Matrices and Data Frames 

To subset a matrix or data frame, a similar bracket notation is used. However, because matrices and data frames are two-dimensional structures, two arguments are required in the brackets. The first argument references the row index and the second input references the column index.

Note: subsetting data frames is very common when analyzing data, as data are usually stored in data frames.

The examples below are applied to a data frame. Subsetting matrices work in the same way.

```{r}
mult_data

## Subsetting the value in the first row and second column
mult_data[1,2]

## Subsetting the values in the first two rows and first two columns
mult_data[1:2,1:2]

## Subsetting the values in rows 1, 4 and columns 2, 3
mult_data[c(1,4),2:3]

## Subsetting the entire first row.
## Comma as the 2nd argument means want all columns
mult_data[1,]

## Subsetting the entire second column
## Comma as the 1st argument means want all rows
mult_data[,2]

## Subsetting all but the first row
mult_data[-1,]

## Subsetting all but the third column
mult_data[,-3]

## Subsetting the third column twice
mult_data[,c(3,3)]

## Subsetting columns 1, 2, 3 and duplicating the third column
mult_data[,c(1:3,3)]
```

#### Subsetting Lists

A list is subset to any of its elements by using double brackets, `[[ ]]`.

```{r}
mult_list[[3]]
```
Double brackets are limited to one value. In order to subset multiple elements of a list, single brackets can be used. 

```{r}
mult_list[2:3]
```
### Subsetting with Labels

#### Label Names with Data Structures

Take a look at the output for `mult_data`. Notice there is an extra row on top (as if they are headings or labels for the columns) and an extra column (as if they are labels for the rows).

```{r}
mult_data
```

Descriptive names can be added to columns, rows, or values within data structures. For data frames and lists, column names can be added while creating the data structure.

```{r}
mult_data2 <- data.frame(Factor=char_vect, Fives=x, Threes=y)
mult_data2
```

```{r}
mult_list2 <- list(Factor.Table=mult_data, FiveThree=z, 
                   Logical.values=log_vect, DaysPerWeek=7)
mult_list2
```

Column, row, or value labels can also be added to an existing data structure using one or more of the following functions: `names()`, `rownames()`, and `colnames()`. The appropriate function(s) to use depends on the type of data structure needing label names. See the examples below:

```{r}
x ## For comparison
## Labeling values in a vector
names(x) <- char_vect
x
```
```{r}
## Labeling rows and columns in a matrix
mult_mat2 <- cbind(x,y)
rownames(mult_mat2) <- char_vect
colnames(mult_mat2) <- c("Fives","Threes")
mult_mat2
```

```{r}
## Labeling columns in a data frame
colnames(mult_data) <- c("Factor","Fives","Threes")
mult_data
```
```{r}
## Labeling values in a list
names(mult_list) <- c("Factor.Table", "FiveThree", "Logical.values", 
                      "DaysPerWeek")
mult_list
```

The `names()` function can also be used to display the names of a data structure. 

```{r}
names(x)
names(mult_data)
names(mult_list)
```

#### Subsetting with Labels

Data structures with labels can be subsetted using their labels. The name of the column or element can be used within the brackets or after the `$` symbol.

```{r}
mult_data[,"Fives"]
mult_data$Fives
```

```{r}
mult_list[["Logical.values"]]
mult_list$Logical.values
```
### Subsetting with Logical Expressions

#### Logical Expressions

It may be useful to use a logical expression to determine which, if any, elements in a vector have certain qualities. When a logical  expression is applied to a vector, the result will be a vector containing TRUE where the condition is satisfied and a FALSE where the condition is not satisfied. 

```{r}
## Are any elements greater than 15?
x > 15

## Are any elements greater than or equal to 15?
x >= 15

## Are any elements exactly equal to 15?
x == 15

## Are any elements not equal to 15?
x != 15
```

Multiple logical expressions can be combined using either and or or.

```{r}
## Are any elements in x greater than 15 AND the same element in y is equal to 12?
x > 15 & y == 12

## Are any elements greater than 15 OR less than 7?
x > 15 | x < 7
```

#### Subsetting with Logical Expressions

The logical vectors that result from logical expression statements can be used in the brackets to subset a vector just to the values that satisfy the condition.    

```{r}
x[x>15]
```

Instead of yielding a logical vector for each entry in a vector, the `which()` function yields the positions in the vector where the condition is satisfied. `The which()` function can also be used in the brackets to subset a vector to just the values that satisfy the condition. 

```{r}
which(x>15)
x[which(x>15)]
```

Two-dimensional data frames can be subsetting similarly, using the an argument for the row and another for the column.

## Loading Data Sets in R

We will cover two main ways to load data sets in R:

1. Data sets from R.
2. Data set from a local file.

### Data Sets from R

Base R comes pre-loaded with some data sets. You can view them by typing `data()`. You can use the specific data set by typing out its name in the list. To view the `airquality` data set:

```{r}
airquality
```

Notice, however, that this data frame does not appear in the environment. Personally, I like to assign such data frames to an object so it appears on my environment. 

```{r}
Data.air <- airquality
```

Data sets could also come from an R package. For example, the `Davis` data set comes from the `carData` package. So install (if needed) and load the package, and type the name of the data set. Again, I prefer to assign the data set to an object so it appears on the environment.

```{r}
library(carData)
Data.Davis <- carData::Davis
```

Notice that I typed `carData::Davis`. This command tells R to go into the `carData` package and use the data set called `Davis`. 

You could have typed `Davis` as well. R would have searched all loaded packages and find the data set called `Davis`. This would work fine if there is no other data set or function with the same name. If another package has a data set or function with the same name, there will be a conflict, and R will use the package that was most recently loaded. This is why I recommend typing the name of the package as well. 

### Data Sets from Local File

A file containing a data set can be easily loaded into the workspace. Be sure the file is in the working directory (see Section \@ref(wd) on how to set the working directory).

If the data file is comma separated (`.csv`), the function `read.csv()` is used to read in the data.

```{r eval=FALSE}
read.csv("Data1.csv", header=TRUE)
```

The argument `header=TRUE` tells R that the first row of the file is a label for the columns (header). If the file does not have a header, specify `header=FALSE`. 

Another common type of data file is a `.txt` file. In this case, use `read.table()`.

```{r eval=FALSE}
## file has headers
read.table("Data1.txt", header=TRUE) 
```






