# Working with Data Structures {#subset}

Imagine you are working on a data frame in R. For most analyses, common tasks will include subsetting the data frame, performing math operations on the data frame, and computing summaries of some columns of the data frame. You will learn how to perform these tasks in this section. 

We will create the same data structures in section \@ref(structures) of the notes to use as examples in this section.

```{r}
x <- c(5,10,15,20,25)
y <- c(3,6,9,12,15)
z <- c(x,y)
log_vect <- c(TRUE, FALSE, T, F)
char_vect <- c("ByOne", "ByTwo", "ByThree", "ByFour", "ByFive")
mult_mat <- cbind(char_vect,x,y)
mult_data <- data.frame(char_vect,x,y)
mult_list <- list(mult_data, z, log_vect, 7)
```

## Subsetting Data Structures

Once a data structure is created and stored, it may be useful in subsequent analyses to subset that data structure. You will learn how to subset data structures using base R functions in this section. Similar operations can be performed using functions from the `dplyr` package, and you will learn these in a later section. 

### Subsetting with Brackets

#### Subsetting Vectors

The typical notation to subset vectors is a set of brackets, `[ ]`, used directly after the variable name of the vector. The information given inside the brackets specifies which elements (in terms of index) should be extracted from the vector. 

The position index of any vector begins at 1 and uses consecutive positive integers up to the length of the vector. For example, to obtain the 3rd element of the vector `x`:

```{r}
x[3]
x ## for comparison
```

Note: Indexing in R starts from 1. R is designed for data analysis, statistics, and mathematics, where standard notations begin counting at 1 (e.g. statisticians think in terms of sample size). This is unlike indexing from 0 for other software such as Python and C, who audiences tend to be programmers and software engineers.

Multiple positions can be specified by using a vector in the brackets.

```{r}
x[c(1,3)] ##1st and 3rd elements
x[2:3] ##2nd and 3rd elements
```

If a position is duplicated in the brackets, the corresponding value in the vector will be subset twice. 

```{r}
x[c(1,1)] ##1st elements displayed twice
```

If a negative sign is used in the brackets, the specified value(s) will be omitted instead of included in the subset.

```{r}
x[-1] ##omit element 1
x[-(2:3)] ##omit elements 2, 3
x[-c(1,4)] ##omit elements 1, 4
```

Instead of specifying the positions that you would like included or excluded from the subset, you can specify a logical vector in the brackets. R will include the elements corresponding to TRUE and exclude the elements corresponding to FALSE.  

```{r}
x[c(F,T,T,F,T)]
```

If the number of logical inputs does not match the length of the vector, R will repeat the logical inputs until there is a logical input corresponding to each vector element.

```{r}
x[c(F,T)]
```

The resulting subset vectors will only be stored if you define a variable name and assign the subset vector to it.

```{r}
evens <- x[c(F,T)]
x
evens
```

The brackets following a vector can also be used to replace or add one or more values in a vector.

```{r}
x2 <- x

x2[6] <- 30 ## add 30 as element 6
x2

x2[7:9] <- c(35,40,45) ##then add 3 values
x2

x2[c(1,3)] <- 0 ##change value of elements 1, 3 to 0
x2
```

#### Subsetting Matrices and Data Frames 

To subset a matrix or data frame, a similar bracket notation is used. However, because matrices and data frames are two-dimensional structures, two arguments are required in the brackets. The first argument references the row index and the second input references the column index.

Note: subsetting data frames is very common when analyzing data, as data are usually stored in data frames.

The examples below are applied to a data frame. Subsetting matrices work in the same way.

```{r}
mult_data

## Subsetting the value in the first row and second column
mult_data[1,2]

## Subsetting the values in the first two rows and first two columns
mult_data[1:2,1:2]

## Subsetting the values in rows 1, 4 and columns 2, 3
mult_data[c(1,4),2:3]

## Subsetting the entire first row.
## Comma as the 2nd argument means want all columns
mult_data[1,]

## Subsetting the entire second column
## Comma as the 1st argument means want all rows
mult_data[,2]

## Subsetting all but the first row
mult_data[-1,]

## Subsetting all but the third column
mult_data[,-3]

## Subsetting the third column twice
mult_data[,c(3,3)]

## Subsetting columns 1, 2, 3 and duplicating the third column
mult_data[,c(1:3,3)]
```

#### Subsetting Lists

A list is subset to any of its elements by using double brackets, `[[ ]]`.

```{r}
mult_list[[3]]
```
Double brackets are limited to one value. In order to subset multiple elements of a list, single brackets can be used. 

```{r}
mult_list[2:3]
```
### Subsetting with Labels

#### Label Names with Data Structures

Take a look at the output for `mult_data`. Notice there is an extra row on top (as if they are headings or labels for the columns) and an extra column (as if they are labels for the rows).

```{r}
mult_data
```

**Descriptive names** can be added to columns, rows, or values within data structures. For data frames and lists, column names can be added while creating the data structure.

```{r}
mult_data2 <- data.frame(Factor=char_vect, Fives=x, Threes=y)
mult_data2
```

```{r}
mult_list2 <- list(Factor.Table=mult_data, FiveThree=z, 
                   Logical.values=log_vect, DaysPerWeek=7)
mult_list2
```

Column, row, or value labels can also be added to an existing data structure using one or more of the following functions: `names()`, `rownames()`, and `colnames()`. The appropriate function(s) to use depends on the type of data structure needing label names. This information is summarized in Table \@ref(tab:labels) below:

Table: (\#tab:labels) Labels for R Data Structures

| Data Structure | `names()`  | `colnames()` | `rownames()` | 
|:----------:|:-------------:|:------:|:------:|
| Vector |  Elements | -  |  - |
| List |  List items   | - | - | 
| Matrix | - | Column names | Row names |
| Data frame | Column names | Column names | Row names |

My advice: use `names()` for one-dimensional data structures, `colnames()` and `rownames()` for two-dimensional data structures. Examples are displayed below:

```{r}
x ## For comparison
## Labeling values in a vector
names(x) <- char_vect
x
```
```{r}
## Labeling rows and columns in a matrix
mult_mat2 <- cbind(x,y)
rownames(mult_mat2) <- char_vect
colnames(mult_mat2) <- c("Fives","Threes")
mult_mat2
```

```{r}
## Labeling columns in a data frame
colnames(mult_data) <- c("Factor","Fives","Threes")
mult_data
```
```{r}
## Labeling values in a list
names(mult_list) <- c("Factor.Table", "FiveThree", "Logical.values", 
                      "DaysPerWeek")
mult_list
```

The `names()` function can also be used to display the names of a data structure. 

```{r}
names(x)
names(mult_data)
names(mult_list)
```

#### Subsetting with Labels

Data structures with labels can be subsetted using their labels. The name of the column or element can be used within the brackets or after the `$` symbol.

```{r}
mult_data[,"Fives"]
mult_data$Fives
```

```{r}
mult_list[["Logical.values"]]
mult_list$Logical.values
```
### Subsetting with Logical Expressions

#### Logical Expressions

It may be useful to use a **logical expression** to determine which, if any, elements in a vector have certain qualities. When a logical  expression is applied to a vector, the result will be a vector containing TRUE where the condition is satisfied and a FALSE where the condition is not satisfied. 

```{r}
## Are any elements greater than 15?
x > 15

## Are any elements greater than or equal to 15?
x >= 15

## Are any elements exactly equal to 15?
x == 15

## Are any elements not equal to 15?
x != 15
```

Multiple logical expressions can be combined using either **and** or **or**.

```{r}
## Are any elements in x greater than 15 AND the same element in y is equal to 12?
x > 15 & y == 12

## Are any elements greater than 15 OR less than 7?
x > 15 | x < 7
```

#### Subsetting with Logical Expressions

The logical vectors that result from logical expression statements can be used in the brackets to subset a vector just to the values that satisfy the condition.    

```{r}
x[x>15]
```

Instead of yielding a logical vector for each entry in a vector, the `which()` function yields the indexes in the vector where the condition is satisfied. `The which()` function can also be used in the brackets to subset a vector to just the values that satisfy the condition. 

```{r}
which(x>15)
x[which(x>15)]
```

Two-dimensional data frames can be subsetting similarly, using the an argument for the row and another for the column.

## Loading Data Sets in R

We will cover two main ways to load data sets in R:

1. Data sets from R.
2. Data set from a local file.

### Data Sets from R

Base R comes pre-loaded with some data sets. You can view them by typing `data()`. You can use the specific data set by typing out its name in the list. To view the `airquality` data set:

```{r}
head(airquality)
```

You can use the `head()` function to print out the first 6 rows of a data frame instead of displaying the entire data frame. This gives the reader a good idea of the data without cluttering the document. 

Notice that this data frame does not appear in the environment. Personally, I like to assign such data frames to an object so it appears on my environment. 

```{r}
Data.air <- airquality
```

Data sets could also come from an R package. For example, the `Davis` data set comes from the `carData` package. So install (if needed) and load the package, and type the name of the data set. Again, I prefer to assign the data set to an object so it appears on the environment.

```{r}
library(carData)
Data.Davis <- carData::Davis
```

Notice that I typed `carData::Davis`. This command tells R to go into the `carData` package and use the data set called `Davis`. 

You could have typed `Davis` as well. R would have searched all loaded packages and find the data set called `Davis`. This would work fine if there is no other data set or function with the same name. If another package has a data set or function with the same name, there will be a conflict, and R will use the package that was most recently loaded. This is why I recommend typing the name of the package as well. 

### Data Sets from Local File

A file containing a data set can be easily loaded into the workspace. Be sure the file is in the working directory (see Section \@ref(wd) on how to set the working directory).

If the data file is comma separated (`.csv`), the function `read.csv()` is used to read in the data.

```{r eval=FALSE}
read.csv("Data1.csv", header=TRUE)
```

The argument `header=TRUE` tells R that the first row of the file is a label for the columns (header). If the file does not have a header, specify `header=FALSE`. 

Another common type of data file is a `.txt` file. In this case, use `read.table()`.

```{r eval=FALSE}
## file has headers
read.table("Data1.txt", header=TRUE) 
```

## Math Operations on Data Structures

The standard arithmetic operators can be applied to data structures  containing numeric values. To perform the same constant arithmetic operation on each element of a numeric data structure, you will use the object name, the arithmetic operation symbol, and the constant. 

```{r}
x <- unname(x) ## remove the name for x
x
x*2
x/5
x^3
```

You may have related numeric information stored in two data structures of the same length / dimensions and you may want to perform an arithmetic operation on each pair of related values. 

```{r}
x
y
x+y
x*y
y/x
```

## Applying Functions on Data Structures

Functions can be applied to data structures, for example

```{r}
sum(x)
mean(mult_mat2)
```

When working with a data frame, we often want to compute some numerical summaries of some or all of the columns. The `apply()` function applies a pre-existing or user-defined function to the data structure in a manner specified by the user. In this unit, we will just cover the most basic application of the `apply()` function. Variants of this function and user-defined functions will be covered later.

Suppose you are working on the `airquality` data set, that is stored in the object `Data.air`. The first four columns represents different numeric characteristics associated with air quality and weather conditions. You want to find the average value in each column.

The `apply()` function is used for a two-dimensional data structure. It uses three arguments: data, margin, and function. The data argument specifies the matrix or data frame to which the function should be applied, margin specifies whether you want the function applied to the rows, columns, or individual values, and function specifies the function to be applied.   

```{r}
## 2 since we want to apply the mean to the columns
## use 1 if want to apply mean to the row
## applying to columns is more common as each column 
## in a data frame represents a characteristic
apply(Data.air[,1:4], 2, mean)
```

Notice the NA for two of the columns? This happens when at least one entry in those columns has a missing value. Missing values in R are denoted by `NA`. If you look at the data frame, you can find some NAs in these columns. 

If you want R to ignore values and exclude them in calculating the mean, add an argument `na.rm=TRUE` in the `apply()` function.

```{r}
apply(Data.air[,1:4], 2, mean, na.rm=TRUE)
```
Suppose you want to count the number of missing values in the column `Ozone`. The `is.na()` function returns a logical identifying whether an entry is missing or not. 

```{r}
## the head function returns only the first 6 entries of a vector
## or first 6 rows of a data frame
head(is.na(Data.air$Ozone))
```

Recall that `TRUE` and `FALSE` are considered to be 1 and 0 respectively. So you can find the number of missing values for `Ozone` with

```{r}
sum(is.na(Data.air$Ozone))
```
