# Functions {#func}

## Motivation

Imagine you are using R to work on a project. You may notice that you have copied and pasted blocks of code in a number of places of your code, with minor tweaks in the value of a particular variable. You will likely be able to replace these blocks of code with a function. A **function** is a reusable block of code designed to perform a specific task. Copying and pasting blocks of code in a number of places is sometimes called **hard-coding**. A benefit of using functions is reducing copying and pasting errors. Manually changing the value of a variable across different blocks of code often leads to mistakes.

You have been using functions that come from base R, or come with other packages. You will now learn how create your own functions.

Functions are defined with three main components: 

- the function **name**, 
- the **arguments** that the function will take, 
- and the **body**, which is the code that should be executed when the function is called.

## Basic Syntax {#basicFunc}

Suppose you want to convert temperature from Celsius to Fahrenheit in your code. You can create a function in the following manner:

```{r}
CtoF <- function(celsius) {
  celsius * (9/5) + 32
}
```

In this particular example, the function name is `CtoF()`. It only takes in one argument named `celsius`. The value of the argument is the value that can change each time the function is used. We then use an open brace, press enter to enter the body of the function, and then press enter and then use a close brace. The body of the function are the commands that are executed when the function is used. 

To use this function to convert 27 degrees Celsius to Fahrenheit

```{r}
CtoF(27)
```

In this example, you can think of the following sequence of events happening when executing the function named `CtoF(27)`:

- The value of `27` is assigned to `celsius`,
- the math operations as written in the body are performed on `celsius`,
- the value after the operations is returned from the function,
- the value of `27` is no longer stored in `celsius`. In fact, `celsius` is not stored in the environment.

Note: when creating your own function, it is best to avoid using naming your function or arguments with words that are associated with functions in base R.

## Returning a Value

A function in R returns the value of its last evaluated expression. You could explicitly use `return()` in the body to clearly state what you want returned. The function below performs identically to `CtoF()`:

```{r}
## explicit return
CtoFexp <- function(celsius) {
  fahrenheit <- celsius * (9/5) + 32
  return(fahrenheit)
}

CtoFexp(27)
```

In R, `return()` is not needed when returning a single object, which is unlike other programming languages. If returning multiple objects, `return()` should be used, see Section \@ref(funcMult).

## Arguments and Default Values

A function can take more than one argument (or even zero arguments such as `getwd()`), and any number of arguments can have a **default** value. Suppose we want to convert Celsius to Fahrenheit, but also want to function to round to the nearest whole number, by default.

```{r}
CtoFrounded <- function(celsius, decimals = 0) {
  fahrenheit <- celsius * (9/5) + 32
  round(fahrenheit, decimals)
}

CtoFrounded(27)
CtoFrounded(celsius = 27, decimals = 1)
```

In the function `CtoFrounded()`, a second argument appears, `decimals`, and is set to 0 by default with the equal sign. This argument does not need to be entered if you want to use the default value when using the function. The first argument `celsius` is not set to any default, and must be specified when calling the function.

You can also use `CtoFrounded(27, 1)` to get 27 degrees Celsius converted to 1 decimal place. The values for the arguments entered in the function must be in the same order as the order of the arguments when the function is created. 

```{r}
CtoFrounded(27, 1)
```

Alternatively, you can use `CtoFrounded(decimals = 1, celsius = 27)`. The order does not matter if you specify the names of the arguments.

```{r}
CtoFrounded(decimals = 1, celsius = 27)
```

You can also decide to have no defaults for any argument at all

```{r}
CtoFnodef <- function(celsius, decimals) {
  fahrenheit <- celsius * (9/5) + 32
  round(fahrenheit, decimals)
}

CtoFnodef(27, 1)
```

## Argument Validation {#validation}

Ideally, your function should check if the argument entered satisfies certain conditions, and outputs either an error message if the argument does not work, or a warning message if the argument is suspicious. This is called **argument validation**. Consider our function `CtoF()`. To convert Celsius to Fahrenheit, the argument entered must be numeric, and the numeric should be within a reasonable range for temperatures.

```{r error=TRUE}
CtoFcheck <- function(celsius) {
  if (!is.numeric(celsius)) {
    stop("celsius must be numeric ")
  }
  if (celsius < -100 | celsius > 100){
    warning("Is celsius within acceptable range?")
  }
  celsius * (9/5) + 32
}

CtoFcheck("blah")
CtoFcheck(250)
```

The `stop()` function stops execution, while `warning()` will print out the warning message but still executes the code and prints out the result. 

Notice `if()` statements are used here. Basically, an `if()` statement checks whether a certain condition is TRUE, if so, the code in the curly braces get executed. If FALSE, the code in the curly braces is skipped, and code outside the curly braces get executed. We will cover `if()` statements in more detail in in Section \@ref(control).

When I write functions, I normally add the parts regarding argument validation last. I would test that my function works with valid arguments, then add the argument validation parts, then test with invalid arguments. 

## Return Multiple Objects {#funcMult}

R functions return a single object. If you need to return multiple objects, you can return those objects into a named list. For example,

```{r}
# Returning multiple items using a list
get_summary_stats <- function(vec) {
  my_mean <- mean(vec)
  my_sd   <- sd(vec)
  
  # Return objects as a list and name them
  return(list(mean = my_mean, sd = my_sd)) 
}

get_summary_stats(c(4,3,1,2,8))
get_summary_stats(c(4,3,1,2,8))$mean
```

## Scope

Variables inside a function are **local**. This means that the value of a variable is not stored in the environment after the function is run (see Section \@ref(basicFunc) where I described the sequence of events when the function is executed). This means that if there is a variable with the same name outside the function, the variable outside is not overwritten. For example,

```{r}
a <- 3 ## global

toy1 <- function(){
  a <- 10 ## local
  return(a + 20)
}

toy1() ## function uses local a

a ## global untouched
```

This means that we do not have to be concerned with names of variables inside a function that we create, as these variables will not overwrite variables with the same name outside the function. 

