# Simulations {#sims}

## Motivation

When we are analyzing data, there is always going to be some degree of uncertainty, as there is **randomness** in a lot of phenomena that we observe in our world. An example that is common in statistics comes from the field of sample surveys. A poll is conducted to see what proportion of voters will vote each candidate in an upcoming election. Even with a well-designed poll and a large sample size, the value of the sample proportion that will vote for a candidate has some randomness to it. If we get another random sample, the sample proportion is likely to be different (but hopefully close). The act of obtaining a random sample, or **sampling**, from a population introduces some randomness. 

Statisticians and data scientists often want to investigate the degree of randomness involved in what is being studied. Sometimes, this can be done mathematically, but oftentimes, the mathematics is extremely challenging. You may also realize in the previous paragraph that one way of studying the randomness in polls is to repeatedly obtain random samples. But it is practically unfeasible to repeatedly go out in the field to collect data.

This is why simulations are used. We do not need to solve complex mathematics, and we can simulate the studies many times (in the thousands, at least) with computers, saving time and money than if we physically ran these studies many times.

You will see how sampling introduces randomness and then learn how to create simulations to quantify the degree of randomness or uncertainty. 

## Sampling

The `sample()` function in R randomly draws a selected number of items from a list of items (the list is usually a vector), and the probabilities of each item being selected are the same for all items. A simple example of randomness is tossing a fair coin a set number of times (equal chance of landing heads or tails).

```{r}
fairCoin <- c("heads", "tails")
sample(fairCoin, size = 5, replace = TRUE)
```
The `sample()` function has three main arguments:

- A vector of values from which to sample.
- How many values from the vector to draw.
- If you want to sample with or without replacement.

**Sampling with replacement** means that when we sample a value from the vector, the value gets returned to the vector and can be sampled again.

**Sampling without replacement** means that means that when we sample a value from the vector, the value does not get returned to the vector and cannot be sampled again.

Run the code in the block above a few times. You will notice that the returned vector is not always the same, so the result is unpredictable. This is unlike most of the earlier code you have seen in the course notes. 

Another use of sampling is in model validation (i.e. assessing model accuracy). The original data set is randomly split into two mutually exclusive data sets: one called training and the other called test. The training set is used to build the model, and model assessment is done on the test set. Suppose we want to split the `Davis` data set from the `carData` package.

```{r}
library(carData)
Data.Davis <- carData::Davis
n <- dim(Data.Davis)[1]
## when dim() is applied to a dataframe, two values get returned
## the num of rows and number of columns. The number of rows is 
## equal to the sample size

trainInd <- sample(1:n, size = ceiling(n/2), replace = FALSE)
## each observation belongs to a row. So randomly select half of
## row indexes without replacement so the two sets are mutually ## exclusive

## split data into training and test
trainData <- Data.Davis[trainInd,]
testData <- Data.Davis[-trainInd,]
```

You have also seen sampling before in Section \@ref(distributions), where functions such as `rnorm()` and `rbinom()` draw random numbers from known probability distributions such as the normal and binomial distributions. Try typing `rnorm(10)` a few times. You should get a vector containing 10 random draws from a standard normal, and the vectors are not always the same.

## Replicate

You have seen how sampling is inherently random. In order to quantify the uncertainty introduced through sampling, we want to be able to repeat, or **replicate**, the sampling a lot of times. 

The `replicate()` function repeatedly evaluates a given function or expression, returning a matrix where each column contains the output of each replicate. There are two arguments: the number of replications and the function to be replicated. 

Going back to the coin tossing example, suppose we want to toss a fair coin 5 times, and replicate this experiment for 6 replicates.

```{r}
replicate(6, sample(fairCoin, size = 5, replace = TRUE))
```

From section \@ref(control), you may think of using a for loop to execute these steps. A for loop could do the same, but the code is longer. 

## Simulation

According to @exploring, a simulation consists of three main steps:

1. Decide what to simulate. What process do we want to imitate?
2. Figure out how to simulate one value.
3. Replicate step 2 many times.

In the previous block of code, we see these steps in action:

- We want to imitate the results from tossing a fair coin 5 times.
- We use the `sample()` function to simulate one "value", which is a vector storing the outcomes from 5 tosses of the coin.
- We use the `replicate()` to repeat the previous step for a total of 6 replicates. 

### Worked Example: Sampling Distribution of the Sample Mean {#ch6sim}

In introductory statistics, you learned that if $X_i \sim N(\mu, \sigma^2)$, then $\bar{X} \sim N(\mu, \frac{\sigma^2}{n})$ (where $\bar{X} = \frac{X_1 + X_2 + \cdots + X_n}{n}$ and $X_i$ are independent. Independence is achieved via random sampling). This result is sometimes called the sampling distribution of the sample mean, and can be proved mathematically. However, such proofs are not the scope of this class. 

This means that if we have data from a $N(\mu, \sigma^2)$ distribution, then the distribution of the sample means from a large number of random samples each of size $n$ is $N(\mu, \frac{\sigma^2}{n})$. We will see how we can use simulation to illustrate the sampling distribution of the sample mean. 

Suppose the data follow a standard normal distribution, i.e. $X \sim N(0,1)$. The sampling distribution of the sample mean tells us that based on samples each of size 10, $\bar{X}_{10} \sim N(0, \frac{1}{10})$. So the three main steps of this simulation would be:

1. We want to simulate 10 random draws from a standard normal distribution, compute their sample mean, and store this value.
2. Use `rnorm(10)` to simulate 10 draws from a standard normal distribution, then compute the sample mean of these 10 draws, and store this value.
3. Repeat step 2 for a large number of times, say for a total of 10,000 replicates.

Step 3 involves using the `replicate()` function, where we supply the number of replicates and a function that gives the result from step 2, i.e. the sample mean from 10 random draws from a standard normal. While there is no function in base R that does this, we can easily write our own function to do this (see Section \@ref(func) on how to write functions)! 

```{r}
sampMean <- function(number){
  mean(rnorm(number))
}
```

Our created `sampMean()` computes the sample mean of a certain number of draws (that is supplied) from a standard normal distribution. `sampMean()` executes step 2.

For step 3, we can just use `replicate()` on `sampMean()`. Suppose we want 10,000 replicates. We store the value of the sample mean from 10,000 random samples, each with size 10. 

```{r}
Reps <- 10000
storeSampMeans <- replicate(Reps, sampMean(10))
```

Intuitively, we know the sample means from each sample are different. We can print out a few of these values to confirm this idea.

```{r}
head(storeSampMeans)
```
If the sampling distribution of the sample mean is correct, we know that $\bar{X}_{10} \sim N(0, \frac{1}{10})$, theoretically. Our simulation should be consistent with this theory. 

We can create a density plot of the 10,000 sample means to visualize the uncertainty from sampling. We can see in figure \@ref(fig:ch6plot) that most of the sample means are close to 0, and values further from 0 are less likely to occur. This also looks like a normal distribution. 

```{r ch6plot, fig.cap='Density Plot of Sample Means from 10,000 Random Samples of Size 10', fig.align='center'}
plot(density(storeSampMeans), main="")
```

We know that the mean and variance of the 10,000 sample means should be close to 0 and 0.1 theoretically. 

```{r}
mean(storeSampMeans)
var(storeSampMeans)
```

The variance measures how spread out the sample means are, and is considered a measure of the uncertainty associated with the sample mean from samples of size 10.

Pay attention to the difference between the number of replicates, which is 10,000, and the sample size (number of data points drawn) of each replicate, which is 10. People often confuse these two. 

We have just created a simulation that imitates the sampling distribution of the sample mean when the data are standard normal and each sample has size 10!

Hopefully, this example convinces you that with a large number of replicates, results from a simulation match very closely with theory.

Note: you could have gotten similar results by using a for loop, using the code below. However, the for loop is a bit longer as we need to pre-allocate, and is less efficient. 

```{r eval=FALSE}
## Repeat for total 10,000 times
Reps <- 10000 
## pre allocate vector to store the 10,000 sample means
sampleMeansLoop <- c(NA, Reps) 

for (i in 1:Reps){
  ## draw 10 numbers from standard normal, find the mean, and store
  sampleMeansLoop[i] <- mean(rnorm(10))
}

plot(density(sampleMeansLoop), main="")
mean(sampleMeansLoop)
var(sampleMeansLoop)
```

In many other situations, mathematical theory can be difficult to prove. For example, what is the sampling distribution of the median from a non-normal population? While finding this result will be challenging mathematically, we can use simulation to get very close to the result. 

### Considerations with Simulations

In the example above in Section \@ref(ch6sim), we compared results from a simulation with the corresponding theoretical values, so we could see that simulations work. However, if we do not know the theoretical values, two questions will come to mind:

- How many replicates do we need? We only know the values from the simulation converge (get close) to the theoretical values as we increase the number of replicates. Using more replicates will make the simulation run longer on your computer.

- Related to the previous question, how close is close enough? How do you know your value from the simulation is close enough to the theoretical value? There is no way of knowing if the theoretical value is unknown.

## Random Seeds

Some of the functions introduced in this section (for example `sample()` and `rnorm()`) **generate random numbers**. Evaluating the same command multiple times will yield different results. The long-run properties are somewhat predictable (e.g. the sample mean when drawing a large number of random numbers from `rnorm()` will be close to the true mean of the normal distribution), but individual outcomes are not predictable. For example, if we run the command `rnorm(1)` twice, we get two different numbers

```{r}
rnorm(1)
rnorm(1)
```

To stay consistent with principles of reproducible research as mentioned in Section \@ref(reproduce), you may need to reproduce the exact results from a simulation. To generate the same random values each time a command is run, the random generation seed can be set by using the `set.seed()` function. There is one required argument, the seed number. 

Seeds operate in a defined order, so once the seed is set, the next random generation commands will use the next seeds in that order. Thus, setting the seed once at the beginning of a program will give the same random values each time the program is run, but each command within the program will generate different results. To yield the same values each time a random generation command is run, the `set.seed()` function must be rerun prior to each command. 

```{r}
## With a set seed specified once
set.seed(5)
rnorm(3)
rnorm(3)

## With a set seed specified each time
set.seed(5)
rnorm(3)
set.seed(5)
rnorm(3)
```

If you copy and paste the code of the simulation from the worked example in Section \@ref(ch6sim), you will realize that the density plot and the sample mean and variance of the 10,000 sample means are close, but not exactly the same. This is because I did not set the seed. The overall observation from the simulation would be similar regardless.

Being able to reproduce the same exact results is likely to be required if you want to get your work published. 

You may want to consider using `set.seed()` in homework questions that involve generating random numbers, so that you can reproduce your result if needed. Be careful if you clear your environment:

- Running the command `rm(list = ls())` does not remove the seed.
- Clicking the broom icon removes the seed. 


