# Apply Family {#applyfam}

## Overview

We often want to apply the same function separately to each element of a vector. R has the `apply` family of functions that implement loops in specialized use cases. This family of functions are useful since it:

- is compact,
- can iterate over different data structures and return collated results, and
- use anonymous functions to improve performance.

This family of functions include:

- `apply()`: Apply a function over the margins of an array.
- `tapply()`: Apply a function over subsets of a vector.
- `lappy()`: Loop over a list and evaluate a function on each item.
- `sapply()`: Similar to `lapply()` but attempt to simplify the result.

## `apply()`

You learned about the `apply()` function in Section \@ref(funcDF). It is to apply a function over the margins of an array. To reiterate: it uses three arguments: data, margin, and function. The data argument specifies the matrix or data frame to which the function should be applied, margin specifies whether you want the function applied to the rows, columns, or individual values, and function specifies the function to be applied. For example, to find the average weight and height all respondents in the `Davis` data set from the `carData` package:

```{r}
library(carData)
Data.Davis <- carData::Davis
apply(Data.Davis[,2:3], 2, mean)
##2 for columns, 1 for rows
```
You may realize that one could use a loop to obtain these values. The `apply()` function is more succinct, although it does not really run faster than a loop [@peng2015].

We so often compute sums or means of columns or rows that there functions that produce the same output as equivalent `apply()` versions:

- `rowSums(X)` equivalent to `apply(X, 1, sum)`.
- `rowMeans(X)` equivalent to `apply(X, 1, mean)`.
- `colSums(X)` equivalent to `apply(X, 2, sum)`.
- `colMeans(X)` equivalent to `apply(X, 2, mean)`.

These functions are faster than `apply()` especially when `X` is large [@peng2015].

## `tapply()`

The `tapply()` function is used for a vector of values that can be categorized by a second vector of values. It uses three arguments: data, categorizations, and function. The data input specifies the vector of values that should be grouped and to which the function should be applied, categorizations specifies the category of each corresponding value in the data vector, and function specifies the function to be applied to each resulting group. For example, to find the average weight of female and male respondents in the `Davis` data set.

```{r}
tapply(Davis$weight, Davis$sex, mean)
```


## `lapply()` and `sapply()`

The `lapply()` and `sapply()` functions are both used for lists. Once the specified function is applied, `lapply()` returns a list and sapply() attempts to simply the result. They use two arguments: data and function. The data argument specifies the list of elements to which the function should be applied and function specifies the function. If the function has other arguments, they can be specified as additional arguments in `lapply()` and `sapply()`. For example:

```{r}
v1 <- list(a = 1:5, b = rnorm(10))
lapply(v1, mean)

##get the mean of 1:5 and the mean for 10 draws from standard normal
```

You can use `lapply()` to evaluate a function multiple times each with a different argument. This can be useful if you want to vary the value of an argument in a simulation. For example, to replicate drawing $n$ values from a standard uniform distribution, with $n = 1, 2, 3, 4$.

```{r}
v2 <- 1:4
lapply(v2, runif)
```

When you pass a function to `lapply()`, `lapply()` takes elements of the list and passes them as the first argument of the function you are applying. In the above example, the first argument of `runif()` is `n`, the number of random draws, and so the elements of the sequence `1:4` are all passed to the `n` argument of `runif()`.

Functions that you pass to `lapply()` may have other arguments. For example, the `runif()` function has a `min` and `max` argument. When these are unspecified, the default values are 0 and 1 respectively, which gives the standard uniform distribution. If these arguments need to be specified, they are passed to `lapply()` next. For example, to replicate drawing $n$ values from a uniform distribution with `min=3` and `max=10`, with $n = 1, 2, 3, 4$.

```{r}
lapply(v2, runif, min = 3, max = 10)
```
You can also pass your own created function to `lapply()`

**Anonymous functions** can also be passed to `lapply()` (and other functions in the apply family). Anonymous functions are functions you create that have no variable name and or not stored in the environment. For example, you have a couple of matrices and you want to extract the first row of each one.

```{r}
twoMats <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))

twoMats

lapply(twoMats, function(extract) { extract[1,] })
```

I recommend using anonymous functions only if the function is simple and if you are going to only use it once. Otherwise, create a function and name it before `lappy()`, and pass the function to `lapply()`. For example

```{r}
row1 <- function(extract){
  extract[1,]
}

lapply(twoMats, row1)
```
Notice that the function `row1` now appears in the environment. 

## `sapply()`

`sapply()` is similar to `lapply()`, but `sapply()` attempts to simplify the result of `lapply()`.

- If the result is a list where every element is length 1, then a vector is returned.
- If the result is a list where every element is a vector of the same length and greater than 1, a matrix is returned, where each column represents each vector.
- Otherwise, a list is returned, just like `lapply()`. 

Look at the output from the following, and compare with the output from `lapply()`. 

```{r}
sapply(v1, mean)
sapply(twoMats, row1)
```
I find working with vectors and matrices easier than working with lists, so I prefer using `sapply()`. 

