# Data Wrangling (dplyr) {#dplyr}

## Motivation

The data structure that we will be dealing with most often will be data frames. When we read data in to R, they are typically stored as a data frame. A data frame can be viewed like an EXCEL spreadsheet, where data are stored in rows and columns. As you saw in Section \@ref(ggplot), we want the data frame to have this basic structure:

+ Each row of the data frame corresponds to an observation.
+ Each column of the data frame corresponds to a variable.

Sometimes, data are not structured in this way, and we will have to transform the data to take on this basic data structure. This process is called data wrangling. The most common and basic operations to transform the data are:


+ Select a subset of rows of the data frame based on some criteria.
+ Rearrange rows in some order.
+ Select a subset of columns of the data frame.
+ Create new columns based on existing variables.
+ Change column names.
+ Change the order of columns.

You may realize that you can already perform some of these operations in Sections \@ref(Rbasics) and \@ref(subset). In this section, you will learning how to perform these operations using another core package from the `tidyverse` called `dplyr`, which has gained popularity. Its advocates say the syntax associated with `dplyr` is more readable for the human eye, although its detractors will say that `dplyr` introduces many function names its syntax strays from base R. 

This section is heavily based on [R for Data Science](https://r4ds.hadley.nz/) [@wickham2023], Chapter 3. We will be covering the basics of data wrangling with `dplyr`. Feel free to consult the book for a more thorough explanation of how these functions work. I have added some of my own commentary.

### Motivating Example

For this section, we will use a data set on flights that departed from New York City in 2013. It comes from the `nycflights13` package. It contains all 336,776 flights that departed from New York City in 2013 and comes from the US Bureau of Transportation Statistics

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
Data <- nycflights13::flights
```

### dplyr Basics

`dplyr` functions typically work in the following manner:

1. Specify data frame.
2. Select which rows and columns to work on.
3. Output data frame.

One of the key operators that we will use is called the **pipe**, `\>`. Pipes pass the result of one function straight into the next function, which makes code clean, linear, and easy to read from top to bottom like a sentence. The easiest way to interpret a pipe is "and then". For example

```{r}
Data |>
  filter(dest == "IAH") |> 
  group_by(month) |> 
  summarize(arr_delay = mean(arr_delay, na.rm = TRUE))
```

What this block of code does is:

1. Takes the data frame called `Data`, and then
2. Filters the observations with destination being `IAH`, and then
3. Groups the observations by month, and then
4. Computes the mean arrival delay for each month.

Compare with the code using base R without pipes:

```{r}
tapply(Data[Data$dest == "IAH",]$arr_delay, Data[Data$dest == "IAH",]$month, mean, na.rm = TRUE)
```
People tend to like the linearity and flow of pipes.


## Operations on Rows

### Select Rows Based on Condition

The `filter()` function from the `dplyr` package selects rows of a data frame based on a condition. For example, to select flights that arrival delays of more than 120 minutes:

```{r}
Data |> 
  filter(dep_delay > 120)
```

You can use an and operator, `&`, for multiple conditions such as arrival delays more than 120 minutes that departed before 1200:

```{r}
Data |> 
  filter(dep_delay > 120 & dep_time < 1200)
```
You can use an or operator, `|`, for example, to find flights in January and February:

```{r}
Data |> 
  filter(month == 1 | month == 2)
```

The `%in%` operator can be used in place of `|` and `==`. It keeps rows where the variable equals one of the values on the right:

```{r}
Data |> 
  filter(month %in% c(1, 2))
```

Notice in these examples that the resulting data frame is printed but not stored. Remember to assign the output to a variable name if you wish to store it. For example, to store the flights on New Year's day:

```{r}
jan1 <- Data |> 
  filter(month == 1 & day == 1)
```

### Arrange Rows in Order

The `arrange()` function arranges the rows based on the value in columns. Specifying multiple columns will arrange by the first column, then by the second, and so on. The default order is small to large. To arrange by month, then by day, and then by departure time:

```{r}
Data |> 
  arrange(month, day, dep_time)
```

To change the order of just the departure times:

```{r}
Data |> 
  arrange(month, day, desc(dep_time))
```

### Counting

To count the number of occurrences of each origin and destination combination:

```{r}
Data |>
  count(origin, dest, sort = TRUE)
```
`sort = TRUE` sorts from largest to smallest.


## Operations on Columns

When operating on columns, we usually want to

- Create new columns (variables) based on existing columns.
- Select specific columns.
- Change the names of columns.
- Change the position of the columns.

### Create New Columns

You can create new columns based on existing columns using `mutate()`. For example, to create a new column called `speed` in miles per hour:

```{r}
Data |> 
  mutate(
    speed = distance / air_time * 60
  )
```

By default, the new column is placed as the last column. You can add `.before = 1` within `mutate()` to place it before column 1.

```{r}
Data |> 
  mutate(
    speed = distance / air_time * 60,
    .before = 1
  )
```
To place it after column 1:

```{r}
Data |> 
  mutate(
    speed = distance / air_time * 60,
    .after = 1
  )
```

Alternatively, using `.keep` within `mutate()` will only keep the columns that were involved in `mutate()`:

```{r}
Data |> 
  mutate(
    speed = distance / air_time * 60,
    .keep = "used"
  )
```

### Select Columns

With data frames with many columns, you may want to only focus on a subset of the columns. To select only the departure and arrival delays:

```{r}
Data |> 
  select(dep_delay, arr_delay)
```

There are a number of helper functions you can use within `select()`. These can be useful if there is a pattern in the naming of the columns:

- `starts_with("abc")`: matches names that begin with “abc”.
- `ends_with("xyz")`: matches names that end with “xyz”.
- `contains("ijk")`: matches names that contain “ijk”.
- `num_range("x", 1:3)`: matches `x1`, `x2` and `x3`.

You can rename variables as you `select()` them. 

```{r}
Data |> 
  select(tail_num = tailnum)
```

The new name appears on the left-hand side of the `=`, and the old variable appears on the right-hand side.

### Rename Columns

If you want to keep all the existing variables and just want to rename a few, you can use `rename()`:

```{r}
flights |> 
  rename(depTime = dep_time)
```

### Relocate Columns

You can use `relocate()` to move variables around. You may want to place important ones in front, or group related variables together. BY default, the variables are moved to the front:

```{r}
Data |> 
  relocate(time_hour, air_time)
```

Or you can specify where to place them after or before:

```{r}
Data |> 
  relocate(time_hour, air_time, .after = day)
```





## Pipes

Pipes are most useful when you want to combine multiple operations. For example, you wanted to find the fastest flights to Houston’s IAH airport while also printing out relevant information:

```{r}
Data |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```

Without pipes:

```{r}
flights1 <- filter(Data, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))
```

Notice multiple objects are created in the intermediate steps.

The keyboard shortcut for a pipe is Ctrl + Shift + M. You may have to change this in RStudio options, via Tools, Global Options, Code and check the box "Use native pipe operator". There is another pipe operator `%>%` which comes from the `magrittr` package. `|>` is part of base R. 

## Grouping {#grouping}

You can group your data frame to set it up for easier analysis. We often want to summarize data across groups. For example, to find the median arrival delay by carrier:

```{r}
Data |>
  group_by(carrier) |> 
  summarize(med_arr_delay = median(arr_delay, na.rm = TRUE))
```

You can also use `n()` to find the number of observations in each group

```{r}
Data |> 
  group_by(carrier) |> 
  summarize(med_arr_delay = median(arr_delay, na.rm = TRUE), 
    n = n()
  )
```
You can also summarize by more than 1 group. For example, to find the number of flights for each combination of departure airports and arrival airports:

```{r}
Data |> 
  group_by(origin, dest) |> 
  summarize(n = n()
  )
```
Note: if you realize that the functions in Section \@ref(grouping) are similar to `tapply()` and `table()`, you are correct!

### Select Rows within Group

There are five handy functions that allow you to extract specific rows within each group:

- `df |> slice_head(n = 1)` takes the first row from each group.
- `df |> slice_tail(n = 1)` takes the last row in each group.
- `df |> slice_min(x, n = 1)` takes the row with the smallest value of column `x`.
- `df |> slice_max(x, n = 1)` takes the row with the largest value of column `x`.
- `df |> slice_sample(n = 1)` takes one random row.

For example, the following code finds the flights that are most delayed upon arrival at each destination:

```{r}
Data |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```




