# Type I and II Errors, Significance Level, Power {#prac}

## Introduction

Section \@ref(hyp) introduced hypothesis testing for quantitative data (mean, paired means, two independent means). In this section, we will cover some additional concepts associated with hypothesis testing: type I and II errors, significance level, and power. 

## Type I and Type II Errors

The decision we make from hypothesis testing is not going to be perfect, as we have inherent uncertainty in estimators due to random sampling. So we may sometimes reject a null hypothesis when it is true, or we may fail to reject a null hypothesis when the alternative hypothesis is true. These errors are called **type I** and **type II** errors respectively. 

We go back to the university bookstore worked example in Section \@ref(eg1samplet). The null and alternative hypotheses were $H_0: \mu = 101.75, H_a: \mu > 101.75$.

- The type I error in this study is concluding that students pay more than \$101.75, on average, when the average amount paid is \$101.75.

- The type II error in this study is concluding that there is not enough evidence that students pay more than \$101.75, on average, when the average amount paid greater \$101.75.

Next, we discuss how we can measure or control the probability of making these errors. 

## Significance Level and Type I Error

The probability of making a type I error is controlled by the **significance level**, $\alpha$, of the hypothesis test. The significance level is the probability of wrongly rejecting the null hypothesis when the null hypothesis is actually true. So if we conduct a hypothesis test with significance level $\alpha = 0.05$, we are saying that we are willing to have a 5% probability of making a type I error if the null hypothesis is true.

The value of the significance level for a hypothesis should be made in consultation with a subject matter expert and understanding the practical cost of making a type I error. From an ethical standpoint, this value should be decided upon before collecting and evaluating any data, as the significance level should not be adjusted in order to be able to conclude that the null hypothesis is rejected. 

We will run a Monte Carlo simulation to demonstrate the significance level determines the probability of making a type I error. This simulation is very similar to the one that was done in Section \@ref(CP) to explain coverage probabilities. You should realize that the probability of making a type I error is directly related to coverage probability which continues with the idea written in Section \@ref(inference) about how confidence intervals and 2-sided hypothesis tests are consistent. The code below does the following:

- Simulate $X_1, \cdots, X_{10}$ i.i.d. from standard normal.
- Calculate a $t$ statistic for the hypothesis test using equation \@ref(eq:9-teststatMean). 
  - The hypotheses will be $H_0: \mu = 0, H_a: \mu \neq 0$. 
  - Note that our data are simulated from a standard normal, so the null hypothesis is true in this simulation. 
- Assess if the calculated $t$ statistic is more extreme than the critical value.
  - We set the significance level to be $\alpha = 0.05$.
  - The critical value is $t_9^* = 2.262157$ since we are conducting a 2-sided test with $n=10$.
- Repeat these steps for a total of 10 thousand replicates.
- Count the number of times we reject the null hypothesis, and divide by the number of replicates. This value estimates the probability of making a type I error.
  - If we did things correctly the estimated probability of making a type I error should be close to the value of the significance level of 0.05.
  
```{r}
## sig level
alpha <- 0.05

## function that simulates data from standard normal,
## then constructs test for mean, and assesses if test
## wrongly rejects H0
mean_type1 <- function(n, alpha){
  
  X <- rnorm(n) ##draw n data points from standard normal
  
  ##calculate elements needed for CI of mean
  xbar <- mean(X) ##sample mean
  SE <- sd(X)/sqrt(n) ##SE of sample mean
  t.stat <- xbar/SE
  crit <- qt(1-alpha/2, n-1) ##critical value
  
  ## check if reject H0
  reject <- abs(t.stat) > crit
 
  return(reject)
  
}

## write another function to replicate this simulation, and
## find the proportion of samples that wrongly reject H0
reps <- 10000

rep_mean_type1 <- function(n, alpha){
  
  result <- replicate(reps, mean_type1(n, alpha))
  return(mean(result))
}

## should be close to the value of the sig level, alpha, 
## which is 0.05 in this simulation

set.seed(31)
rep_mean_type1(10, alpha)
```

The estimated probability of making a type I error is 0.0531, which matches the significance level which was set to be 0.05. 

**Thought question**: Feel free to play around with the code and adjust the value of the significance level, change the alternative hypothesis to a 1-sided test, or any other factor. The estimated probability of making a type I error should always match the significance level. 

If the type I error of a hypothesis test does not match its significance level, the test is considered unreliable.

## Power and Type II Error {#power}

The **power** of a test is inversely related to the probability of making a type II error. The power of a test is its probability of being able to correctly reject the null alternative hypothesis, when the alternative hypothesis is true. Based on this definition, the power is the complement of the probability of making a type II error, i.e. power of a test is 1 minus the probability of making a type II error. 

Calculating the power of a hypothesis test can be done by hand but is not usually a simple exercise. We will instead use Monte Carlo simulations to estimate the power of a hypothesis test. 

Suppose we want to estimate the power of the hypothesis test for the mean if the data are i.i.d from $N(1,1)$ (and so has a population mean of 1), with sample size of 10, and significance level set to 0.05, and the hypotheses are $H_0: \mu = 0, H_a: \mu \neq 0$. The code below does the following:

- Simulate $X_1, \cdots, X_{10}$ i.i.d. from $N(1,1)$.
- Calculate a $t$ statistic for the hypothesis test using equation \@ref(eq:9-teststatMean). 
  - The hypotheses are $H_0: \mu = 0, H_a: \mu \neq 0$. 
  - Note that our data are simulated from $N(1,1)$, so the alternative hypothesis is true in this simulation. 
- Assess if the calculated $t$ statistic is more extreme than the critical value.
  - We set the significance level to be $\alpha = 0.05$.
  - The critical value is $t_9^* = 2.262157$ since we are conducting a 2-sided test with $n=10$.
- Repeat these steps for a total of 10 thousand replicates.
- Count the number of times we reject the null hypothesis, and divide by the number of replicates. This value estimates the probability of correctly rejecting the null hypothesis, i.e. the power of this test.

```{r}
## sig level
alpha <- 0.05

## function that simulates data from N(1,1),
## then constructs test for mean, and assesses if test
## correctly rejects H0
mean_power <- function(n, alpha){
  
  X <- rnorm(n, mean = 1, sd = 1) ##draw n data points N(1,1)
  ##calculate elements needed for CI of mean
  xbar <- mean(X) ##sample mean
  SE <- sd(X)/sqrt(n) ##SE of sample mean
  t.stat <- xbar/SE
  crit <- qt(1-alpha/2, n-1) ##critical value
  
  ## check if reject H0
  reject <- abs(t.stat) > crit
 
  return(reject)
  
}

## write another function to replicate this simulation, and
## find the proportion of samples that wrongly reject H0
reps <- 10000

rep_mean_power <- function(n, alpha){
  
  result <- replicate(reps, mean_power(n, alpha))
  return(mean(result))
}

## should be close to the value of the sig level, alpha, 
## which is 0.05 in this simulation

set.seed(31)
rep_mean_power(10, alpha)
```

The estimated power of this test is 0.8003. So if the data come from a $N(1,1)$, the power of testing $H_0: \mu = 0, H_a: \mu \neq 0$ using a sample of size 10 and significance level of 0.05 is 0.8003.

Note: we must specify a specific value for the parameter that makes the alternative hypothesis true. In this simulation, we simulated data from a distribution with population mean 1. The value of the power will change if we used another value for the mean of the distribution. This value directly affects the effect size. The power is specific to this value for the alternative hypothesis, as well as to the specific values of $n$ and $\alpha$. 

**Power analysis** or power calculations are often done in well-designed experiments to determine what minimum sample size is needed for a hypothesis test to guarantee the hypothesis test achieves a certain power with a pre-determined significance level and specific alternative hypothesis (which impacts the effect size).

**Thought question**: How would you tweak the code above to do a power analysis to find the minimum sample size needed to have an estimated power of at least 0.90 (assuming every other factor remains the same)?

### Factors Affecting Power

Ideally, we would like the power of our hypothesis test to be large. So what are the factors of the power of a hypothesis test? Recall that power measures the probability of being able to correctly reject a null hypothesis, and that larger values (in magnitude) of a test statistic provide more evidence against the null hypothesis. So the factors that make a test statistic larger (in magnitude) also lead to an increase in the power of a hypothesis test. As mentioned in Section \@ref(factors), these are:

- The effect size. There are a couple of definitions of effect size:

    1. The difference between the true value of the parameter and the value of the parameter under the null hypothesis.
    2. A "standardized" difference between the true value of the parameter and the value of the parameter under the null hypothesis. For a 1-sample t-test, this is $\frac{\mu_x - \mu_0}{s}$. For a 2-sample t-test, this is $\frac{(\mu_x - \mu_y) - (\text{diff in null})}{\sqrt{s_x^2 + s_y^2}}$. General guidelines for interpretating the effect size when defined this way:
        - 0.2: small effect.
        - 0.5: moderate effect.
        - 0.8: large effect.

In either definition, larger effect sizes lead to larger test statistics, in magnitude. This should make sense since a larger effect size implies the sample mean is likely to deviate more from the null value if the null hypothesis is true. 

- The sample size $n$. Larger sample sizes lead to larger magnitudes for the test statistic.

- The standard error of the estimator. Smaller standard errors of the estimator lead to larger magnitudes for the test statistic.

- Another factor that increases the power of a hypothesis test is to raise the significance level $\alpha$. Think of the following extreme example: Suppose we set the significance level to be as low as possible, $\alpha=0$. Then the hypothesis test will always not reject the null hypothesis, and we never make a type I error. But always not rejecting the null hypothesis means that if the alternative hypothesis is true, we will always make a type II error, so the power will be 0. This illustrates the probabilities of making a type I and type II error have an inverse relationship. Reducing one is done at the expense of the other. 

### Worked Examples

We will use a couple of functions from the `pwr` package to compute power. These functions are useful if you are working in terms of standardized effect size. 

1. We go back to the example in Section \@ref(eg1samplet). A university bookstore claims that, on average, a student will pay \$101.75 per class for textbooks. The student newspaper suspects that the claimed average is too low and investigates by randomly selecting 9 courses from the current semester offerings and finding the textbook costs for each. The students conducting the investigation reason that with the collection of courses being offered courses, the textbook prices are normally distributed. 

- What is the effect size based on the sample? 
- What is the power of their planned test to identify a significant difference when the true cost per textbook is moderately higher?
- What is the minimum sample size to attain a power of at least 0.80 with moderate effect size?

From the question, we have $H_0: \mu = 101.75, H_a: \mu > 101.75$.

```{r}
## data
news.samp <- c(140, 125, 150, 87, 170, 195, 94, 127, 53)

## effect size
efs <- (mean(news.samp) - 101.75)/sd(news.samp)
efs

library(pwr)

## power from pwr package
pow <- pwr.t.test(d = 0.5, n = 9, sig.level = 0.05, 
                  type = "one.sample", 
                  alternative = "greater")$power
pow

## Determine the sample size required for power of 0.8
## need to round up to whole number
min.n <- ceiling(pwr.t.test(d = 0.5, sig.level = 0.05, 
                            power = 0.8, type = "one.sample", 
                            alternative = "greater")$n)
min.n
```

The effect size is `r efs`. 

The power with moderate effect size is `r pow`.

The minimum sample size for moderate effect size with at least 0.80 power is `r min.n`.

2. A weight study was conducted on 307 sets of twins in Australia over the age of 50 to determine if there is evidence that identical twins have significantly different weights. What is the power if the effect size is small?    

```{r}

## Calculate the power
pow2 <- pwr.t.test(d = 0.2, n = 307, sig.level = 0.05, 
                   type = "paired", alternative = "two.sided")$power
pow2

## Determine the sample size required for power of 0.8
min.n2 <- ceiling(pwr.t.test(d = 0.2, sig.level = 0.05, 
                     power = 0.8, type = "paired", 
                     alternative = "two.sided")$n)
min.n2

```

The power with small effect size is `r pow2`.

The minimum sample size for small effect size with at least 0.80 power is `r min.n2`.

3. A biologist is interested in whether different species of iris have different sepal widths. He is particularly interested in whether the Versicolor and Virginica species are significantly different. With random samples of 150 flowers from each species, what is the power of the biologist's test when the effect size is small?  

```{r}
## Calculate the power
pow3 <- pwr.t.test(d = 0.2, n = 150, sig.level = 0.05, 
                   type = "two.sample", 
                   alternative = "two.sided")$power
pow3

## Determine the sample size required for power of 0.8
min.n3 <- ceiling(pwr.t.test(d = 0.2, sig.level = 0.05, 
                             power = 0.8, type = "two.sample", 
                             alternative = "two.sided")$n)
min.n3
```

The power with small effect size is `r pow3`.

The minimum sample size for small effect size with at least 0.80 power is `r min.n3`.

Note that the `pwr.t.test()` function assumes equal sample sizes when applied to a 2-sample t-test. The `pwr.2t.test()` function should be used when sample sizes are not the same. 

4. The biologist conducting the experiment to identify  differences in sepal widths between iris species has planted the irises, but 5 of the Versicolor plants have died as have 9 of the virginica plants. What is the power of the biologist's updated test when the effect size is small? 

```{r}
## Calculate the power
pow4 <- pwr.t2n.test(d = 0.2, n1 = 145, n2 = 141, 
                     sig.level = 0.05, 
                     alternative = "two.sided")$power
pow4
```

The power with small effect size is `r pow4`.

5. The biologist decides that it is possible to plant additional virginica irises for the study. How many more should the biologist plant for 50% power? 

```{r}
## Determine the sample size required for power of 0.5
min.n5 <- ceiling(pwr.t2n.test(d = 0.2, n1 = 145, sig.level = 0.05, power = 0.5, alternative = "two.sided")$n2)
min.n5 - 141
```

The biologist needs to plant `r min.n5 - 141` more virginica irises for the study to achieve 0.50 power. 

## Statistical Vs Practical Significance

However, even with a p-value being reported, we need to be careful with conflating a "statistically significant" result (i.e. null hypothesis is rejected) with a "practically significant" result. 

We had earlier mentioned in Section \@ref(power) that the power of a hypothesis test can be increased when the sample size gets larger. This implies that we may be able to reject the null hypothesis when we have a small effect size but we have a huge sample size. 

Consider this hypothetical example. We have $H_0: \mu = 100, H_a: \mu \neq 100$, with $\bar{x} = 101, \sigma=10$, and various values for the sample size $n$. Suppose these values represent IQ scores. We work out the p-values as well as the corresponding 95% confidence intervals for the mean, and present the results in Table \@ref(tab:practical) below. 

Table: (\#tab:practical) Sample Size on P-Values and CIs

| $n$ | p-value | 95% CI |
| :------- | :------- | :------- | 
| 25 | 0.617 | (97.080, 104.920)|
| 350 | 0.061 | (99.952, 102.408) | 
| 400 | 0.455 | (100.020, 101.980) | 
| 2,500 | $5.7 \times 10^{-7}$ | (100.608, 101.392) | 

Most would probably agree that the effect size in this example is small, yet for a sample size of 400 or more, we will reject the null hypothesis. When the sample size is huge, the p-value is extremely small, which we may interpret as having strong evidence against the null hypothesis, so we have "statistical significance". On the other hand, the small effect size tells us that the practical significance may be of little consequence. 

A couple of recommendations when reporting results from a hypothesis test:

- If possible, report the corresponding confidence interval, so the reader can see what is the range of plausible values for the parameter and make their own decision. However, not every hypothesis test has an associated confidence interval; we will see one such class of hypothesis tests in the next subsection.

- Researchers need to establish the minimum effect size for a result to be considered practically significant. From an ethical standpoint, this value should be decided upon before collecting and evaluating any data, as the this value should not be adjusted in order to be able to conclude that the null hypothesis is rejected. 

To remind ourselves, we use hypothesis tests to test for "statistical significance". What we are testing is whether the deviation of our sample estimate from its value if the null hypothesis is true can be attributed to random sampling alone or not. If we reject the null hypothesis, the deviation is unlikely to be attributed to random sampling alone. 

To summarize, hypothesis testing is used to assess "statistical significance", but we use subject matter knowledge to assess "practical significance".
